### 2022-12-28 paper.Rmd
editing

    Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---and is usually approached by maximzing instead of minimizing a clustering objective function. Introducing k-plus, this paper presents an extension of the k-means clustering objective to maximize between-group similarity in anticlustering applications. The k-plus criterion represents between-group similarity as discrepancy in distribution moments (means, variance, higher order moments, and covariances), thereby extending the classical k-means criterion, which only reflects how groups differ in their means. It is shown that the k-plus criterion can be reduced to an augmentation of the input data while the original k-means objective remains unchanged. A computer simulation and practical examples show that k-plus anticlustering achieves high between-group similarity with regard to multiple objectives. In particular, optimizing between-group similarity with regard to variances as well as means comes with almost no cost to either criterion; the k-plus extension is therefore generally preferred over classical k-means anticlustering. Higher order moments and covariance structure can also succesfully be equalized. Example are given on how k-plus anticlustering can be applied to real norming data using the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

documentclass     : "apa6"
classoption       : "doc"
output            : 
  papaja::apa6_pdf
---

@@

Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---which seeks homogeneity within clusters and separation between clusters [@rokach2005; @steinley2006]. Anticlustering has many applications in research psychology [@steinley2006; @brusco2019; @papenberg2020]. Examples include splitting tests into parts of equal difficulty [@gierl2017], assigning students to work groups [@baker2002], and assigning stimuli to different, but parallel, experimental conditions [@lahl2006]. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, anticlustering problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods are characterized by (a) an objective function that quantifies between-group similarity and/or within-group heterogeneity and (b) an algorithm that determines how elements are assigned to groups to maximize the objective function Several anticlustering objective functions use pairwise dissimilarity ratings such as the Euclidean distance as input. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between any elements within the same group [@brusco2019; @gallego2013]. By summing all pairwise dissimilarities of elements within the same group, the diversity reflects the total degree of within-group heterogeneity. High within-group diversity simultaneously ensures high between-group similarity.[^equalsizedgroups] Another important anticlustering criterion is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications requiring high within-group heterogeneity should incorporate both dispersion and diversity, and presented a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

@@

Oftentimes, partitioning applications in psychology are focused on ensuring between-group similarity rather than within-group heterogeneity (even though both goals usually coincide). High similarity between-group similarity is for example desirable when preparing experimental conditions that use different stimulus sets that should however be as similar as possible with regard to response-relevant attributes [@lahl2006]. Rather than relying on pairwise (dis)similarity ratings, stimulus selection tasks are usually conducted on the basis of attribute data [or *features*; @dry2009]. For stimuli in psycholinguistic experiments, for example, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, a distance-based anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the stimuli. A different anticlusting approach employs clustering objectives that can directly be computed on the attribute values. K-means is probably the best-known objective of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other parameters---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random group assignment usually yielded more similar variances. When requiring overall between-group similarity, the negligence of other distribution characteristics than the mean is misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that quantifies between-group similarity as discrepancy with regard to other distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. After formally introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups having minimal differences according to multiple distribution characteristics. Practical examples illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

For the purpose of problem formalization, I adopt the notation @steinley2006 provided in his comprehensive synthesis of half a century of research on k-means. K-means clustering (and anticlustering) is used to partition two-way, two-mode data into groups. That is, $N$ elements each having $P$ attributes are assigned to $K$ groups $(C_1., \ldots, C_K)$, $C_k$ being the set of $n_k$ objects in group $k$. The partitioning is subject to the requirement that each element is assigned to exactly one group, formalized by the following two constraints [@papenberg2020]:

@@

The *k*'th centroid $\overline{\mathbf{x}}^{(k)} = (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})$ is a vector of length $P$, where each entry is the mean value of one of the $P$ attributes, computed across all observations belonging the $k$th cluster $C_k$ ($k = 1, \ldots, K$):

@@

In the following, I present a variation of the $\mathit{SSE}$ that quantifies how the variance of numeric data differs between groups. To motivate the new criterion, we first consider a set of one-dimensional observations $(x_1, \ldots, x_N)$. The extension to the general case of $P$ attributes will be straight forward.

The sample variance of a set of observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

@@

Because the sample variance is basically defined as an average, k-means anticlustering can be employed to equalize the variance across groups, which is the central insight motivating the new k-plus criterion. To obtain an adaption of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, we have to compute the squared deviation between all individual values and the mean of the attribute. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$, where

@@

It is unlikely that an exclusive optimization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is of interest without also considering the $\mathit{SSE}$ (i.e., equalizing means). A combination of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, however, can be employed to simultaneously optimize both objectives, which is a reasonable idea when striving for overall between-group similarity. The most simple approach for such a bicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, both criteria are computed independently from each other and then combined into a single criterion through a weighted sum:

@@

To thoroughly evaluate the k-plus approach to anticlustering, three specific k-plus criteria were implemented in a simulation study: (i) "standard" k-plus anticlustering, minimizing differences with regard to the mean and variance; (ii) k-plus *skew/kurtosis*, minimizing differences with regard to the mean, variance, skew and kurtosis; (iii) k-plus *correlation*, minimizing differences with regard to the mean, variance and covariance structure. To implement all k-plus objectives, the simple unweighted sum outlined above approach was employed. Thus, when optimizing a k-plus criterion, the original data matrix was extended to incorporate the additional criteria. Subsequently, the original k-means criterion $SSE$ was maximized on the basis of the augmented data set.

@@

The simulation was implemented using the statistical programming language R [Version 4.2; @R-base]. The R package `anticlust` [Version 0.6.1; @R-anticlust; @papenberg2020] was used to optimize the anticlustering objectives. The R package `faux` [Version 1.1.0; @R-faux] was used to generate the multivariate normal data sets that were processed during the simulation. The R packages `dplyr` [Version 1.0.7; @R-dplyr], `ggplot2` [Version 3.3.5; @R-ggplot2], `papaja` [Version 0.1.0.9997; @R-papaja], `tidyr` [Version 1.1.3; @R-tidyr] and `DescTools` [Version 0.99.45; @R-desctools] were used for data analysis and result presentation. The simulation study is fully reproducible via code and data that is accessible from the accompanying OSF repository (**link TODO**).

@@

The same local maximum search [Method LCW in @weitz1998] was applied to optimize each of the five competing anticlustering objectives. Building on an initial random assignment of elements to groups, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, for each element, all possible exchanges are simulated and the resulting objective values are stored. After each exchange has been tested, the one exchange is realized that improves the objective the most. No exchange is realized if no improvement is possible. The exchange procedure is repeated for each element. Once the last element has been processed, the procedure restarts at the beginning and once again iterates through all elements, repeating the exchange procedure. The procedure stops as soon as an iteration through the entire data set no longer yielded an improvement, i.e., as soon as a local maximum has been found. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. The simulation employed five repetitions of the local maximum search.

@@

After the five anticlustering methods and the random allocation procedure had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I investigated how well the competing methods were able to create similar groups. To this end, I investigated the discrepancy in means, standard deviations, skewness, kurtosis, and correlation between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$). Quantifying discrepancy in correlations ($\Delta \mathit{Cor}$) followed the same rule, but discrepancies in correlations between groups had to be averaged across pairs of features instead of single features.

@@

@@

# Load required packages
library(dplyr)
library(ggplot2)
library(tidyr)
source("./Simulation_Study/0-functions-generate-data.R")

@@

apa_table(
  tab,
  caption = "Results of the simulation study",
  note = "The global results of the simulation study aggregated across all conditions. Cells contain information about the average between-group discrepancy with regard to means, standard deviations, skewness, kurtosis, and correlations. Lower values indicate less discrepancy, i.e., higher between-group similarity. Each cell value is the result of averaging across 30,000 data points (10,000 data sets $\\times$ $K = 2$, $K = 3$ or $K = 4$).",
  escape = FALSE,
  align = c("r")
)

@@

@@

As Figure 1 shows, with increasing $N$, k-means and the k-plus methods converge on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus term in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time---especially when $N$ increases. However, while k-plus anticlustering is capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria apparently aggravates fulfilling each single criterion. K-plus skew/kurtosis and k-plus correlation performed worse at minimizing discrepancy with regard to standard deviations than standard k-plus, which itself performed slightly worse than k-means anticlustering. Still, every k-plus method did what it was supposed to do: k-plus skew/kurtosis was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation was best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to have performed much better than diversity anticlustering with regard to skewness and---in particular---kurtosis, even though the method has been specifically tailored to these criteria. Figure 1 gives a more fine grained assessment of this observation by splitting the results by $N$: Surprisingly, for small $N$, k-plus skew/kurtosis was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: for low $N$, k-plus seems skew/kurtosis to be preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitates to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

Generally, it should be noted that diversity anticlustering is a good all-arounder method that tends to somewhat addressed all distribution characteristics. For the specific objectives that are addressed by the k-plus methods, however, diversity anticlustering was outperformed. As as side note, apart from k-plus correlation, diversity anticlustering was the only method that equalized correlation structure among groups. To the best of my knowledge, this observation has not been made previously, even though the maximum diversity problem has been widely studied.

@@

This section demonstrates how readers can easily employ k-plus anticlustering using the R package `anticlust` [@papenberg2020], using an interface function called `kplus_anticlustering()`, which is available from version 0.6.3 onward. The package can be installed in the R environment using the `install.packages()` command:

@@

Table 2 shows the results of the anticlustering application by listing the descriptive statistics (means, standard deviations and skewness) for each of the 9 groups and for each of the 3 features. Table 2 also includes results of k-means anticlustering and diversity anticlustering. It is shown that---up to two decimals---k-plus anticlustering perfectly matched all groups with regard to the features' means and standard deviations, and the skewness was also very evenly matched. K-means anticlustering also perfectly matched the mean values, but showed decreased performance with regard to similarity in standard deviations and especially skewness. Diversity anticlustering was well-suited to match means, standard deviations and skew, but was slightly outperformed by k-plus. The latter result is not surprising because---unlike k-plus anticlustering---diversity anticlustering does not directly maximize similarity with regard to these criteria, but instead maximizes within-group heterogeneity.

@@

Table 3 illustrates another advantage of k-plus anticlustering over diversity anticlustering: When creating groups of unequal size, diversity anticlustering tends to increase the spread of the data in the largest group in comparison to the other groups. This is due the diversity criterion being a measure of within-group heterogeneity. Because the diversity criterion is a sum rather than an average, the correspondence between within-group heterogeneity and between-group similarity does not hold for unequal-sized groups. K-plus anticlustering strives for between-group similarity regardless of group size.

Unequal group sizes can be requested in `anticlust` by passing the different group sizes to the argument `K`. The following code was used to equalize means and variances between four groups using the k-plus approach:

@@

This function call relied on the default k-plus settings ensuring similarity with regard to means and variances equalized---but with regard to higher order moments or covariances.

@@

Anticlustering is a partitioning method that aims for high between-group similarity and within-group heterogeneity. In this paper, I presented the k-plus criterion, which focuses on maximizing between-group similarity. K-plus represents between-group similarity as discrepancies in distribution moments between groups: mean, variances, skewness, kurtosis, higher order moments, and even covariances. Depending on the application and the needs of a user, a different subset of these objectives can be included in the anticlustering process. The k-plus criterion is an extension of the classical k-means criterion, which only represents how similar groups are with regard to their means. Interestingly, however, all k-plus variations basically do not change the original k-means criterion, because additional criteria can be incorporated by merely augmenting the input data.

A simulation study and examples on real norming data showed that the k-plus criterion is well suited to maximize between-group similarity with regard to multiple criteria. Given that similarity with regard to variances was achieved with almost no sacrifice to similarity in means, k-plus anticlustering should be a default replacement of classical k-means when striving for between-group similarity.[^binarydata] The results did however show that fulfilling multiple criteria at the same time can be aggravated in small data sets. In practice, users can try out different objectives and see which specification fulfils their needs. A reasonable approach is to start "greedy", e.g. by specifying that multiple distribution moments should equalized. If the results of such a greedy setup are subpar, users can reduce their requirements by opting for fewer criteria (e.g., only the mean and variances). Unlike in statistical analysis, there is no problem with "just trying out" different angles when it comes to anticlustering. Moreover, using the open source R package `anticlust`, the (repeated) application of k-plus anticlustering is easily accessible and free.

@@

>>>>

    Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---and is usually approached by maximzing instead of minimizing a clustering objective function. Introducing k-plus, this paper presents an extension of the k-means clustering objective to maximize between-group similarity in anticlustering applications. The k-plus criterion represents between-group similarity as discrepancy in distribution moments (means, variance, higher order moments, and covariances), while the classical k-means criterion only reflects the degree to which groups differ in their means. It is shown that the k-plus criterion can be reduced to an augmentation of the input data while the original k-means objective remains unchanged. A computer simulation and practical examples show that k-plus anticlustering achieves high between-group similarity with regard to multiple objectives. In particular, optimizing between-group similarity with regard to variances as well as means comes with almost no cost to either criterion; the k-plus extension is therefore generally preferred over classical k-means anticlustering. Higher order moments and covariance structure can also succesfully be equalized. Example are given on how k-plus anticlustering can be applied to real norming data using the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

documentclass     : "apa6"
output            : 
  papaja::apa6_pdf
appendix: "appendix.Rmd"

header-includes:
  - \usepackage{setspace}
---

@@

Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---which seeks homogeneity within clusters and separation between clusters [@rokach2005; @steinley2006]. Anticlustering has many applications in research psychology [@steinley2006; @brusco2019; @papenberg2020]. Examples include splitting tests into parts of equal difficulty [@gierl2017], assigning students to work groups [@baker2002], and assigning stimuli to different, but parallel experimental conditions [@lahl2006]. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, anticlustering problems can be formalized as mathematical optimization problems [e.g., @baker2002; @brusco2019; @spath1986; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods are characterized by (a) an objective function that quantifies between-group similarity and/or within-group heterogeneity, and (b) an algorithm that determines how elements are assigned to groups to maximize the objective function. Several anticlustering objective functions use pairwise dissimilarity ratings such as the Euclidean distance as input. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between elements within the same group [@brusco2019; @gallego2013]. By summing all pairwise dissimilarities within the same group, the diversity reflects the total degree of within-group heterogeneity. For anticlustering, high within-group heterogeneity is desired, which simultaneously ensures high between-group similarity.[^equalsizedgroups] Another important anticlustering criterion is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications striving for within-group heterogeneity should incorporate both dispersion and diversity, and they presented a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

@@

Oftentimes, partitioning applications in psychology are focused on ensuring between-group similarity rather than within-group heterogeneity (even though both goals oftentimes coincide). High between-group similarity is for example desirable when preparing experimental conditions that use different stimulus sets, that should however be as similar as possible with regard to response-relevant attributes [@lahl2006]. Rather than relying on pairwise (dis)similarity ratings, stimulus selection tasks are usually conducted on the basis of attribute data. That is, individual stimuli are numerically coded on relevant dimensions [@dry2009]. For stimuli in psycholinguistic experiments, for example, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, a distance-based anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the stimuli. A different anticlusting approach relies on clustering objectives that can directly be computed using the attribute values. K-means is probably the best-known objective of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized, usually using the k-means heuristic [@jain2010; @steinley2006; @brusco2006branch]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other distribution characteristics---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random group assignment usually resulted in more similar variances. When requiring overall between-group similarity, the negligence of all distribution characteristics other than the mean is misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that quantifies between-group similarity as discrepancy with regard to several distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. After formally introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups having minimal differences according to multiple distribution characteristics. Practical examples illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

For the purpose of problem formalization, I adopt the notation @steinley2006 provided in his comprehensive synthesis of half a century of research on k-means. K-means anticlustering is used to partition $N$ elements each having $P$ attributes into $K$ groups $(C_1, \ldots, C_K)$, where $C_k$ is the set of $n_k$ objects in the $k$'th group. The partitioning is subject to the requirement that each element is assigned to exactly one group, formalized by the following two constraints [@papenberg2020]:

@@

The *k*'th centroid $\overline{\mathbf{x}}^{(k)} = (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})$ is a vector of length $P$, where each entry is the mean value of one of the $P$ attributes, computed across all observations belonging the $k$'th cluster $C_k$ ($k = 1, \ldots, K$):

@@

In the following, I present a variation of the $\mathit{SSE}$ that quantifies how the variance of numeric data differs between groups. To motivate the new criterion, we first consider a set of one-dimensional observations $(x_1, \ldots, x_N)$. The sample variance of a set of observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

@@

Because the sample variance is basically defined as an average, k-means anticlustering can be employed to equalize the variance across groups, which is the central insight motivating the new k-plus criterion. Therefore, to obtain an adaption of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, we have to compute the squared deviation between all individual values and the mean of an attribute. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$, where

@@

It is unlikely that an exclusive maximization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is of interest without also considering the $\mathit{SSE}$ (i.e., equalizing means). A combination of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, however, can be employed to simultaneously optimize both objectives, which is a reasonable idea when striving for overall between-group similarity. The most simple approach for such a bicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, both criteria are computed independently from each other and are then combined into a single criterion through a weighted sum:

@@

To thoroughly evaluate the k-plus approach to anticlustering, three specific k-plus criteria were implemented in a simulation study: (i) "standard" k-plus anticlustering, minimizing differences with regard to the mean and variance; (ii) k-plus *skew/kurtosis*, minimizing differences with regard to the mean, variance, skew and kurtosis; (iii) k-plus *correlation*, minimizing differences with regard to the mean, variance and covariance structure. To implement all k-plus objectives, the simple unweighted sum outlined above approach was employed. Thus, when optimizing a k-plus criterion, the original data matrix was extended to incorporate the additional criteria. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set.

@@

The simulation was implemented using the statistical programming language R [Version 4.2, @R-base]. The R package `anticlust` [Version 0.6.1, @R-anticlust; @papenberg2020] was used to optimize the anticlustering objectives. The R package `faux` [Version 1.1.0, @R-faux] was used to generate the multivariate normal data sets that were processed during the simulation. The R packages `dplyr` [Version 1.0.7, @R-dplyr], `ggplot2` [Version 3.3.5, @R-ggplot2], `papaja` [Version 0.1.0.9997, @R-papaja], `tidyr` [Version 1.1.3, @R-tidyr] and `DescTools` [Version 0.99.45, @R-desctools] were used for data analysis and result presentation. The simulation study is fully reproducible via code and data that is accessible from the accompanying OSF repository (**link TODO**).

@@

The same local maximum search [Method LCW, @weitz1998] was applied to optimize each of the five competing anticlustering objectives. Building on an initial random assignment of elements to groups, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, for each element, all possible exchanges with elements in other clusters are simulated and the resulting objective values are stored. After each exchange has been evaluated, the one exchange is realized that improves the objective the most. No exchange is realized if no improvement is possible. The exchange procedure is repeated for each element. Once the last element has been processed, the procedure restarts at the beginning and once again iterates through all elements, repeating the exchange procedure. The procedure stops as soon as an iteration through the entire data set no longer yielded an improvement, i.e., as soon as a local maximum has been found. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. The simulation study always employed five repetitions of the local maximum search.

@@

After the five anticlustering methods and the random allocation procedure had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I investigated how well the competing methods were able to create similar groups. To this end, I investigated the discrepancy in means, standard deviations, skewness, kurtosis, and correlation between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$). Quantifying discrepancy in correlations ($\Delta \mathit{Cor}$) followed the same rule, but discrepancies in correlations had to be averaged across pairs of features instead of single features.

@@

\begin{singlespace}

@@

# Load required packages
library(dplyr)
library(ggplot2)
library(tidyr)

@@

foo <- apa_table(
  tab,
  caption = "Results of the simulation study",
  note = "The global results of the simulation study aggregated across all conditions. Cells contain information about the average between-group discrepancy with regard to means, standard deviations, skewness, kurtosis, and correlations. Lower values indicate less discrepancy, i.e., higher between-group similarity. Each cell value is the result of averaging across 30,000 data points (10,000 data sets $\\times$ $K = 2$, $K = 3$ or $K = 4$).",
  escape = FALSE,
  align = c("r")
)

@@

\end{singlespace}

@@

Figure 1 shows that with increasing $N$, k-means and the k-plus methods converged on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus term in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time---especially when $N$ increases. However, while k-plus anticlustering is capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria apparently aggravates fulfilling each single criterion. Hence, k-plus skew/kurtosis and k-plus correlation performed worse at minimizing $\Delta \mathit{SD}$ than standard k-plus. Still, every k-plus method did what it was supposed to do: k-plus skew/kurtosis was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation was best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to have performed much better than diversity anticlustering with regard to skewness and---in particular---kurtosis, even though the method has been specifically tailored to these criteria. Figure 1 gives a more fine grained assessment of this observation by splitting the results by $N$: Surprisingly, for small $N$, k-plus skew/kurtosis was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: for low $N$, k-plus seems skew/kurtosis to be preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

Generally, it should be noted that diversity anticlustering is a good all-arounder method that tends to address all distribution characteristics. For the specific objectives that are addressed by the k-plus methods, however, diversity anticlustering was outperformed. As as side note, apart from k-plus correlation, diversity anticlustering was the only method that equalized correlation structure among groups. To the best of my knowledge, this observation has not been made previously, even though the maximum diversity problem has been widely studied.

The appendix includes additional figures that split the simulation results according to the other variables that varied between data sets (number of features; correlation between features; standard deviation of features; number of groups). Here, the only meaningful moderating effect on anticlustering performance was that k-plus correlation was more strongly impaired than other methods when the number of features increased. Because the number of covariances increases quadratically when the number of features increases, it gets comparatively more difficult to fulfill the correlation objective with a higher number of features.

@@

This section demonstrates how readers can easily employ k-plus anticlustering using the R package `anticlust` [@papenberg2020] using an interface function called `kplus_anticlustering()`, which is available from version 0.6.3 onward. The package can be installed in the R environment using the `install.packages()` command:

@@

Table 2 shows the results of the anticlustering application by listing the descriptive statistics (means, standard deviations and skewness) for each of the 9 groups and for each of the 3 features. Table 2 also includes results of k-means anticlustering and diversity anticlustering. It is shown that---up to two decimals---k-plus anticlustering perfectly matched all groups with regard to the features' means and standard deviations, and the skewness was also very evenly matched. K-means anticlustering also perfectly matched the mean values, but showed decreased performance with regard to similarity in standard deviations skewness in particular. Diversity anticlustering was well-suited to match means, standard deviations and skew, but was slightly outperformed by k-plus. The latter result is not surprising because---unlike k-plus anticlustering---diversity anticlustering does not directly maximize similarity with regard to these criteria, but instead maximizes within-group heterogeneity.

@@

Table 3 illustrates another advantage of k-plus anticlustering over diversity anticlustering: When creating groups of unequal size, diversity anticlustering tends to increase the spread of the data in the largest group in comparison to the other groups. K-plus anticlustering strives for between-group similarity regardless of group size. Unequal group sizes can be requested in `anticlust` by passing the different group sizes to the argument `K`. The following code was used to equalize means and variances between four groups using the k-plus approach:

@@

This function call relied on the default k-plus setting that ensures similarity with regard to means and variances, but does not consider higher order moments or covariances.

@@

Anticlustering is a partitioning method that aims for high between-group similarity and within-group heterogeneity. In this paper, I presented the k-plus criterion for anticlustering, which focuses on maximizing between-group similarity. K-plus represents between-group similarity as discrepancies in distribution moments between groups: mean, variances, skewness, kurtosis, higher order moments, and even covariances. Depending on the application and the needs of a user, a different subset of these objectives can be included in the anticlustering process. The k-plus criterion is an extension of the classical k-means criterion, which only reflects how similar groups are with regard to their means. Interestingly, however, all k-plus variations basically do not change the original k-means criterion, because additional criteria can be incorporated by augmenting the input data.

A simulation study and examples on real norming data showed that the k-plus criterion is well suited to maximize between-group similarity with regard to multiple criteria. Given that similarity with regard to variances was achieved with almost no sacrifice to similarity in means, k-plus anticlustering should be considered as the default replacement of classical k-means when striving for between-group similarity.[^binarydata] The results did however show that fulfilling multiple criteria at the same time can be aggravated in small data sets. In practice, users can try out different objectives and see which specification fulfils their needs. A reasonable approach is to start "greedy", e.g. by specifying that multiple distribution moments should equalized. If the results of such a greedy setup are subpar, users can reduce their requirements by opting for fewer criteria (e.g., only the mean and variance). Unlike in statistical analysis, there is no problem with "just trying out" different angles when it comes to anticlustering. Moreover, using the open source R package `anticlust`, the (repeated) application of k-plus anticlustering is easily accessible and free.

@@

# (APPENDIX) Appendix {-}

```{r child = "appendix.Rmd"}
```
